Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism

### 2023-11-01 paper/main.qmd
Implement Lars's paper comments

---
title: "Bayesian regression facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism, 
          multiomics integration, ordinary differential equations,
          regulatory anlaysis
abstract: >
  This paper presents Maud, a command-line application that implements Bayesian
  statistical inference for kinetic models of biochemical metabolic reaction
  networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Our paper
  reviews the existing options in this area, presents a case study illustrating
  how Maud can be used to analyse a metabolic network and explains the
  biological, statistical and computational design decisions underpinning Maud.
number-sections: true
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Nicholas Luke Cowie"
    affiliations: 
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
    attributes:
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Lars Keld Nielsen"
    affiliations: 
      - name: The University of Queensland
        department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
        city: St Lucia
        country: Australia
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

In equation \eqref{eq-1}, C represents a vector of metabolite concentrations,
$v$ is a vector of reaction rates, and $\mu$ is the growth rate. The second term
represents the dilution due to cell growth.

In a kinetic model, the rates, $v$, are expressed as a function of the enzyme
concentrations, $E$, the metabolite concentrations, $C$, and a set of parameters,
$\theta$ as shown in equation \eqref{eq-2}

@@

The parameters must include sufficient boundary concentrations and fluxes to
solve equation \eqref{eq-1}.

@@

In a fermentation context, equation \eqref{eq-3} captures the rapid kinetics
inside the cell, while another set of ODEs would be used to describe the
external substrate and product concentrations, which could act as boundary
parameters to equation \eqref{eq-3}.

In the context of kinetic modelling, Bayesian inference is appealing because
it allows uncertainty to be represented appropriately without sacrificing
mechanistic accuracy. Measurement uncertainty can naturally be represented in a
Bayesian measurement model, whereas the prior model can represent quantitative
uncertainty about kinetic parameters. Finally, kinetic rate laws can be
represented in Bayesian data generation models with arbitrarily high fidelity.
See @gelmanBayesianDataAnalysis2020a for more about Bayesian inference and
@gelmanBayesianWorkflow2020 for a discussion of practical Bayesian workflow.

Another advantage is that Bayesian inference problems are well-posed even when
not all parameters are strongly identified. Sloppy models in which measurable
quantities are sensitive to combinations of parameters but not to individual
marginal parameter values are ubiquitous in models of biological systems
[@gutenkunst_2007; @white_2016]. The parameter correlation structure represents
the set of potential models that describe the observed data. As we demonstrate
in @sec-laplace below, capturing this correlation structure is difficult outside
a fully Bayesian context.

@@

The most popular algorithm for fitting Bayesian statistical models is
Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow exploration of
high- dimensional posterior distributions, have robust failure diagnostics
[@vehtariRankNormalizationFoldingLocalization2021] and can incorporate fast
numerical solvers, thereby making inference feasible for Bayesian kinetic
models. Nonetheless, the kinetic modelling literature reports an aversion
to MCMC, rooted mainly in concerns about sampling time and the presumed
difficulty of implementing the required statistical model [@raue_joining_2013;
@saa_construction_2016]. We are only aware of two recent attempts to implement
a Bayesian kinetic modelling approach using MCMC. @stapor_pesto_2018 fitted
detailed kinetic models using relatively inefficient MCMC algorithms that
do not scale well to high dimensional parameter spaces limiting the scope
of modelling. Conversely, @st.johnBayesianInferenceMetabolic2018 utilises an
efficient sampling algorithm but uses approximate kinetics, namely lin-log
kinetics @visser_dynamic_2003, limiting the scope of interpreting parameters and
inferring cellular behaviour in experimental conditions outside the reference
dataset.

There have also been efforts to implement Bayesian inference for kinetic models
without the use of MCMC. Examples of alternative inference methods include
variational inference [@st.johnBayesianInferenceMetabolic2018], rejection
sampling and approximate Bayesian computation [@saa_construction_2016] and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori parameter
configuration [@liebermeister_model_2021; @gopalakrishnan_k-fit_2020;
@stapor_pesto_2018; @raue_joining_2013]. Non-MCMC-based Bayesian kinetic
models have limited utility because they lack reliable diagnostic tools for
verifying that their results approximate the target posterior distribution.
This is a problem because realistic kinetic models tend to induce highly
correlated, non-Gaussian, joint probability distributions [@gutenkunst_2007;
@stapor_pesto_2018].

@@

We constructed a kinetic model of the methionine cycle in Maud’s format using
the description in @korendyaseva_allosteric_2008. The ordinary differential
equation system describing this model is shown in Equation \eqref{eq-meth-ode}.

@@

We used these parameters to simulate steady states in a range of plausible
experimental conditions, again using @saa_construction_2016 as a starting point.
These steady states were then used to generate simulated measurements using the
measurement model described below in @sec-methionine-measurement-model.

### Measurement model {#sec-methionine-measurement-model}

Generating an artificial dataset required a specification of the true
measurement error. For enzyme and metabolite concentration measurements
we specified a standard deviation of 0.1 on natural logarithmic scale,
corresponding to approximately 10% measurement error. For each reaction
measurement a measurement standard deviation of approximately 10% of the
simulated value.

@@

For our main model run, we assumed that all metabolite and enzyme concentrations
were measured, and that there was a reaction measurement for each of the
network’s elementary flux modes.

### Training/validation split

The split between training and validation conditions was selected to achieve a
large difference in fluxes between the two categories. The split was determined
as we are interested in showing how our model can fit to varied conditions, and
conditions closer to the training set are likely to be predicted well without
necessarily learning the system.

### Additional dataset with missing measurements

### Maud input specification

@@

@@

## Findings

### Posterior inference

@@

Analysis of the posterior distributions for kinetic parameters indicated that
these are highly correlated. The marginal posterior distributions for most
kinetic parameters did not shrink significantly compared with the corresponding
marginal prior distributions, even though these parameters’ joint posterior
distribution contained enough information to make accurate out of sample
predictions. In some cases, there were two-dimensional correlations such as
the one shown in the bottom left of @fig-posterior; in this case the
marginal distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to reveal the
underlying correlation structure, as seen in the bottom-right plot in figure
@fig-posterior. This does not mean that the parameters were uncorrelated, but
rather that the correlations involve more than two parameters.

@@

The question naturally arises whether the crucial high-dimensional parameter
correlations are linear or non-linear. This is relevant to the question of
model performance, as linear correlations are easier to correct for. A linearly
correlated posterior space would also be easier to summarise. We address this
question below in @sec-laplace.

This case study illustrates the type of kinetic model and dataset that Maud
can fit. The model we analysed has 10 reactions, 5 state variables and 212
parameters. Generalising from our ability to fit this model in a reasonable time
using Maud, we expect that Maud can be used to fit realistic Bayesian models
of approximately the same size, but not, for example, genome-scale kinetic
models. To fit larger models, faster steady state solving methods or alternative
inference algorithms will be required. @sec-laplace addresses whether Laplace
approximation is a suitable candidate.

@@

Marginal posterior distributions from our main model run. (**Top left**)
Comparison of posterior predictive intervals with simulated flux measurement
values. All the flux measurements are within the predictive intervals,
indicating a good fit. (**Top right**) Comparison of posterior predictive
intervals with simulated concentration measurement values. These also show a
good fit. (**Bottom left**) Pairwise marginal posterior distribution for two
uncorrelated parameters, namely $K_m^{AHC1,hcys-L}$ and $K_m^{AHC1_ahcys}$.
(**Bottom right**) Pairwise marginal posterior distribution for two correlated
parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.

@@

We found that the Laplace method was not able to produce an accurate posterior
approximation for our model and dataset.

Our overall strategy was to compare samples generated using MCMC, which we
were confident could be treated as draws from the true posterior distribution,
with approximate posterior samples generated using the Laplace method. We were
unable to generate approximate posterior samples for our main methionine cycle
case study using Laplace approximation, as the algorithm could not recover from
solver failures caused by unrealistic parameter configurations. We therefore
made a comparison for a simpler model. This input can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>.

MCMC sampling for our comparison input yielded 800 samples in 625 minutes;
Laplace sampling yielded the same number of samples in only one minute.
The diagnostics indicated that our algorithm was able to find the maximum
a posteriori parameter configuration, approximate the Hessian and use these
quantities to generate approximate posterior samples. The results can be found
at <https://github.com/biosustain/Methionine_model/tree/main/results>.

@fig-laplace summarises the results of comparing the samples generated using
each method. As can be seen from the top left plot, the Laplace method does not
provide a good approximation to the true posterior distribution in this case,
as the marginal distribution of the total log probability density is clearly
different. This impression was confirmed using the Kolmogorov-Smirnov test,
which is a test to differentiate two empirical univariate distributions. The
two distributions were significantly different with a p-value indistinguishable
from zero.

The difference between the Laplace approximation output and the true posterior
distribution manifests not only in the parameter space, but also in the
measurement space. The upper right (training data) and lower left (validation
data) plots in @fig-laplace show 5%-95% marginal flux measurement log likelihood
intervals for both methods; lower log likelihood values indicate that the
modelled and measured values are further away. The graph shows that the Laplace
approximation yielded significantly worse predictions than the true posterior,
even for the training data.

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$: see the
bottom right cell of @fig-laplace. This comparison demonstrates that
the Laplace method is not able to capture the correct relationships between
parameters’ distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect that
the Laplace method will produce worse approximations the more complex the target
model. Since the approximation is already unacceptable for our simple test
model, we recommend that Maud users use MCMC sampling in preference to Laplace
approximation if possible when fitting realistic Bayesian kinetic models.

@@

Graphical comparison of approximate posterior samples generated using Laplace
sampling (blue-green) with posterior samples generated using MCMC (dark
grey). (**Top left**) the two sets of samples clearly have different marginal
distributions for the overall log probability variable, indicating that the
Laplace samples do not accurately approximate the target distribution. (**Top
right**) The distribution of marginal posterior predictive log likelihood values
for training data flux measurements shows that the Laplace method tended to
yield much worse predictions compared with the true model. (**Bottom left**)
The Laplace method also tended to produce worse flux predictions for held-back
test measurements. (**Bottom right**) The marginal joint distribution of two
parameters: $K_m^{A,rl}$ and $K_{cat}^{r1}$. The Laplace method is not able
to track the correct joint distribution for this pair of parameters. This is
unsurprising given that the target distribution has position-dependent scales
which are difficult for a linear approximation to capture.

@@

### Effect of missing metabolite concentration measurements

@@

@fig-missing shows that, as might be expected, the model with missing
measurements did not correctly infer the missing ahcys concentrations.
Nonetheless, the remaining measured metabolites were still well predicted,
suggesting that information about the network is still preserved despite the
missing measurements. Comparison of flux measurements in both models also
indicated that removing the ahcys measurement did not result in catastrophic
model failure.

The missing measurements did affect Maud’s ability to infer parameter values
correctly. The lower left plot of @fig-missing shows that the model with full
metabolomics learned the true value for the displayed dissociation constant,
despite this value being far from the mean of the corresponding marginal prior
distribution. In contrast, the model with missing measurements stayed in the
neighbourhood of the prior.

@@

### Application to regulatory understanding

To demonstrate how Maud's output can be used to yield useful metabolic insights
we used the results of our case study to explain why the flux of the enzyme
GNMT is higher in dataset 1 than in dataset 12. GNMT is an irreversible enzyme
that is homotropically activated by its substrate, competitively inhibited by
its product and heterotropically inhibited by 5,10- methylenetetrahydrofolate
(mlthf). The diverse regulation makes it the ideal test case to elucidate
regulatory changes.

@fig-decomposition shows the regulatory description of GNMT, according to the
results our main case study analysis. Each curve shows the marginal posterior
distribution of the ratio of the corresponding regulatory component in dataset
1 compared with dataset 12. A positive value indicates that the component
was increased in dataset 1 relative to dataset 12, with 0 indicating no
difference. The probability, according to our model, of the component acting in
each direction is given by the relative area under the curve on each side of the
zero point.

@@

Illustration of how analysing a system with Maud can yield actionable insights
about the underlying metabolic network.  (**Top Left**) A schematic of the
regulatory interactions associated with the enzyme $GNMT1$. Dashed green
lines represents allosteric activation, dashed red lines indicate allosteric
inhibition and solid red lines represent competitive inhibition. (**Bottom
Left**) Comparison of marginal posterior distributions for $GNMT$ flux in
datasets 1 and 2. (**Right**) Log-scale ratios of the regulatory elements
defined in equation \eqref{eq-decomposition}. Note that the $Reversibility$
and $k_cat$ components are excluded: this is because this reaction was modelled
as irreversible, and $k_{cat}$ was modelled as constant across datasets.
These plots identify why flux in dataset 12 is higher than in dataset 1: the
flux increase is due to allostery and saturation with no control from enzyme
concentration changes.

@@

As well as parameter values, Maud also performs inference for derived quantities
including the components of the regulatory decomposition described below in
equation \eqref{eq-decomposition}, log likelihoods, simulated measurement values
and metabolic control analysis coefficients [@kacser_control_1973].

@@

Maud’s kinetic model decomposes into factors as shown in equation
\eqref{eq-decomposition}.

@@

Metabolite and enzyme measurements are intended to represent the results of
quantitative metabolomics and proteomics experiments. The likelihood functions
for such measurements are shown in equations Equation \eqref{eq-yconc} and
Equation \eqref{eq-yenz}.

@@

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm $\ln$) and known standard deviation
$\sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
consideration that concentrations are constrained to be non-negative, so the
measurement model should avoid assigning positive probability mass to negative
metabolite concentration values. In addition, we expect the precision of most
metabolomics and proteomics experiments to be roughly proportional to the value
of the true measured quantity, which supports a measurement model with constant
coefficient of variation. The measurement standard deviations $\sigma_{𝑐𝑜𝑛𝑐}$
and $\sigma_{𝑒𝑛𝑧}$ are assumed to be known exactly for simplicity; plausible
values can be elicited by considering the likely coefficient of variation of the
measuring apparatus.

Flux measurements, representing the results of quantitative fluxomics analyses,
are modelled using a likelihood function from a standard linear regression
model, as shown in equation \eqref{eq-yflux}. Flux measurements can be obtained
from the results of isotope labelling experiments using metabolic flux analysis,
for example as described in (Young 2014). When entering flux measurements,
it is important only to specify measurements for a network’s free fluxes, as
the values of some steady state fluxes in a metabolic network are constrained
by others, with the result that dependent fluxes cannot typically be measured
separately. If measurements of multiple dependent fluxes are entered,
information will inappropriately be double counted.

@@

Our measurement model improves on analyses of metabolomics and proteomics
data that assume a regression model with normally distributed errors, whether
explicitly using a standard linear model or implicitly using ordinary least
squares fitting. This assumption is undesirable because it implies that the
measured quantity could in principle be negative, and assumes an additive
underlying random process, whereas multiplicative processes tend to better
describe real concentration data.

@@

Maud's parameters are connected with measurable quantities via the steady state
equation \eqref{eq-3}. Posterior sampling using adaptive Hamiltonian Monte
Carlo requires repeatedly solving equation \eqref{eq-3} and finding the
gradients of its solution with respect to all model parameters. This must be
done numerically as analytic solutions are not available for realistic kinetic
models.

@@

To solve equation \eqref{eq-3} and find its gradients, Maud uses
a hybrid method involving two numerical solvers from the SUNDIALS
suite [@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via
their interface from Stan. The hybrid method follows that proposed by
@margossianComputingSteadyStates2018, and involves numerically evolving the ODE
system for a short period of time, then using the difference between the evolved
and starting concentrations as the target for a numerical algebra solver.

@@

This research was funded by the Novo Nordisk foundation using grant number
NNF14OC0009473.

>>>>

---
title: "Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism"
keywords: Bayesian inference, kinetic models of cell metabolism, 
          multiomics integration, ordinary differential equations,
          regulatory anlaysis
abstract: >
  This paper presents Maud, a command-line application that implements Bayesian
  statistical inference for kinetic models of biochemical metabolic reaction
  networks. Maud takes into account quantitative information from omics
  experiments and background knowledge, as well as structural information about
  kinetic mechanisms, regulatory interactions and enzyme knockouts. Our paper
  reviews the existing options in this area, presents a case study illustrating
  how Maud can be used to analyse a metabolic network and explains the
  biological, statistical and computational design decisions underpinning Maud.
number-sections: true
author: 
  - name: "Teddy Groves"
    orcid: 0000-0002-7109-3270
    email: tedgro@biosustain.dtu.dk
    affiliations: 
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
    attributes:
      corresponding: true
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Nicholas Luke Cowie"
    affiliations: 
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
    attributes:
      equal-contributor: true
    note: These authors contributed equally to this research.
  - name: "Lars Keld Nielsen"
    affiliations: 
      - name: The University of Queensland
        department: Australian Institute for Bioengineering and Nanotechnology (AIBN)
        city: St Lucia
        country: Australia
      - id: biosustain
        name: DTU
        department: DTU Biosustain
        city: Kongens Lyngby
        country: Denmark
bibliography: bibliography.bib
format:
  acs-pdf:
    keep-tex: true
    journal:
      id: asbcd6  #This is an identifier for the target journal: from https://pubs.acs.org/, the identifier is the part of the URL after https://pubs.acs.org/journal/, e.g. for J. Org. Chem. it would be joceah.
      type: article  #article, note, communication, review, letter, or perspectivee
      layout: traditional #traditional or twocolumn, twocolumn currently broken for longtables
      hide-email: false # if enabled, hides email addresses (and fax/phone)
  acs-html:
   include-in-header: mathjax.html
   embed-resources: true
  docx: default
---

@@

where C represents a vector of metabolite concentrations, $v$ is a vector of
reaction rates, and $\mu$ is the growth rate. The second term represents the
dilution due to cell growth.

In a kinetic model, the rates, $v$, are expressed as a function of the enzyme
concentrations, $E$, the metabolite concentrations, $C$, and a set of parameters
$\theta$:

@@

The parameters must include sufficient boundary concentrations and fluxes to
solve \eqref{eq-1}.

@@

In a fermentation context, \eqref{eq-3} captures the rapid kinetics
inside the cell, while another set of ODEs would be used to describe the
external substrate and product concentrations, which could act as boundary
parameters to \eqref{eq-3}.

In the context of kinetic modelling, Bayesian inference is appealing because
it allows uncertainty to be represented appropriately without sacrificing
mechanistic accuracy. Measurement uncertainty can naturally be represented in a
Bayesian measurement model, whereas the prior model can represent quantitative
uncertainty about kinetic parameters. Finally, kinetic rate laws can be
represented in Bayesian data generation models with arbitrarily high fidelity.
See @gelmanBayesianDataAnalysis2020a for more details about Bayesian inference
and @gelmanBayesianWorkflow2020 for a discussion of practical Bayesian workflow.

Another advantage is that Bayesian inference problems are well-posed even when
not all parameters are strongly identified. Sloppy parameter models in which
measurable quantities are sensitive to combinations of parameters but not to
individual marginal parameter values are ubiquitous in models of biological
systems [@gutenkunst_2007; @white_2016]. The parameter correlation structure
represents the set of potential models that describe the observed data. As we
demonstrate later, capturing this correlation structure is difficult outside a
fully Bayesian context.

@@

The most popular algorithm for fitting Bayesian statistical models is Markov
Chain Monte Carlo (MCMC). Modern MCMC algorithms enable exploration of
high-dimensional posterior distributions, have robust failure diagnostics
[@vehtariRankNormalizationFoldingLocalization2021] and can incorporate fast
numerical solvers, thereby making inference feasible for Bayesian kinetic
models. Nonetheless, the kinetic modelling literature reports an aversion
to MCMC, rooted mainly in concerns about sampling time and the presumed
difficulty of implementing the required statistical model [@raue_joining_2013;
@saa_construction_2016]. We are only aware of two recent attempts to implement
a Bayesian kinetic modelling approach using MCMC. @stapor_pesto_2018 fitted
detailed kinetic models using relatively inefficient MCMC algorithms that
do not scale well to high dimensional parameter spaces limiting the scope
of modelling. Conversely, @st.johnBayesianInferenceMetabolic2018 utilises an
efficient sampling algorithm but uses approximate kinetics, namely lin-log
kinetics [@visser_dynamic_2003], limiting the scope of interpreting parameters and
inferring cellular behaviour in experimental conditions outside the reference
dataset.

There have also been efforts to implement Bayesian inference for kinetic models
without the use of MCMC. Examples of alternative inference methods include
variational inference [@st.johnBayesianInferenceMetabolic2018], rejection
sampling and approximate Bayesian computation [@saa_construction_2016] and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori parameter
configuration [@liebermeister_model_2021; @gopalakrishnan_k-fit_2020;
@stapor_pesto_2018; @raue_joining_2013]. Non-MCMC-based Bayesian kinetic
models suffer from a lack of reliable diagnostic tools for verifying that their
results approximate the target posterior distribution. This is a problem because
realistic kinetic models tend to induce highly correlated, non-Gaussian, joint
probability distributions [@gutenkunst_2007; @stapor_pesto_2018].

@@

We constructed a kinetic model of the methionine cycle in Maud’s format using
the description in @korendyaseva_allosteric_2008. The ordinary differential
equation system describing this model is

@@

We used these parameters to simulate steady states in a range of plausible
experimental conditions, again using @saa_construction_2016 as a starting point.
These steady states were then used to generate simulated measurements using the
measurement model.

For enzyme and metabolite concentration measurements we specified a standard
deviation of 0.1 on natural logarithmic scale, corresponding to approximately
10% measurement error. For each reaction measurement a measurement standard
deviation of approximately 10% of the simulated value.

@@

For our main model run, we assumed that all metabolite and enzyme concentrations
were measured, and that there was a reaction measurement for each of the
network’s free fluxes.

The simulated experiments and measurements were split into a training and a
validation dataset in a way that achieved a large difference in flux between the
two categories. This was done to evaluate whether the fitted model is able to
extrapolate to conditions well outside the training dataset rather than merely
interpolating between the training data without necessarily learning the system.

@@

## Posterior inference

@@

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared to the
corresponding marginal prior distributions, even though these parameters’
joint posterior distribution contained enough information to make accurate out
of sample predictions. In some cases, there were two-dimensional correlations
such as the one shown in the bottom left of @fig-posterior; in this case the
marginal distribution of the two parameters is roughly banana shaped. More
commonly, however, two-dimensional pair plots were insufficient to reveal the
underlying correlation structure, as seen in the bottom-right plot in figure
@fig-posterior, which shows two marginally independent parameters. This kind of
marginal independence does not mean that the parameters are independent in the
full joint posterior distribution.

@@

The question naturally arises whether the crucial high-dimensional parameter
correlations are linear or non-linear. This is relevant to the question of
model performance, as linear correlations are easier to correct for. A linearly
correlated posterior space would also be easier to summarise. We address this
question in the next section.

@@

Marginal posterior distributions from our main model run. (**Top left**)
Comparison of posterior predictive intervals with simulated flux measurement
values. All the flux measurements are within the predictive intervals,
indicating a good fit. (**Top right**) Comparison of posterior predictive
intervals with simulated concentration measurement values. These also show
a good fit. (**Bottom left**) Pairwise marginal posterior distribution
for two pairwise correlated parameters, namely $K_m^{AHC1,hcys-L}$ and
$K_m^{AHC1_ahcys}$. (**Bottom right**) Pairwise marginal posterior distribution
for two pairwise-uncorrelated parameters, namely $K_m^{MAT3,atp}$ and $K_{cat}^{MAT3}$.

@@

This current case study illustrates the type of kinetic model and dataset that
Maud can fit. The model we analysed has 10 reactions, 5 state variables and 212
parameters. Generalising from our ability to fit this model in a reasonable time
using Maud, we expect that Maud can be used to fit realistic Bayesian models
of approximately the same size, but not, for example, genome-scale kinetic
models. To fit larger models, faster steady state solving methods or alternative
inference algorithms will be required. The Laplace approximation, in which the
Fisher information matrix is used to calculate a normal approximation around
the maximum a posteriori parameter configuration, is a popular strategy for
generating approximate posterior samples while avoiding a full MCMC approach.

Using the recently implemented Laplace approximation in Stan, we were unable to
generate approximate posterior samples for our main methionine cycle case study
using Laplace approximation, as the algorithm could not recover from solver
failures caused by unrealistic parameter configurations. This is not an inherent
issue of the Laplace transformation but highlights some of the challenges around
solving \eqref{eq-3} for real problems.

In order still to assess the potential use of the Laplace approximation, we made
a comparison for a simpler model. This input can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>.
MCMC sampling for this simpler model yielded 800 samples in 625 minutes, while
Laplace sampling yielded the same number of samples in only one minute.
The diagnostics indicated that our algorithm was able to find the maximum
a posteriori parameter configuration, approximate the Hessian and use these
quantities to generate approximate posterior samples. The results can be found
at <https://github.com/biosustain/Methionine_model/tree/main/results>.

Comparing the samples generated using each method shows that the Laplace
approximation does not provide a good approximation to the true posterior
distribution in this case (@fig-laplace). As can be seen from the top left
plot, the total log probability density is clearly different and this was
confirmed by a test of the equality of two empirical univariate distributions
(Kolmogorov-Smirnov, $p=1.7\times10^{-65}$).

The difference between the Laplace approximation output and the true posterior
distribution manifests not only in the parameter space, but also in the
measurement space for both training data (upper right) and validation data
(lower left). The lower log likelihood values indication that the modelled and
measured values are further away and that the Laplace approximation yielded
significantly worse predictions than the true posterior, even for the training
data.

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants $K_{m}^{A,r1}$ and $K_{cat}^{r1}$
(@fig-laplace, bottom right). This comparison demonstrates that the Laplace
method is not able to capture the correct relationships between parameters’
distributions.

This result shows that MCMC, while slower than Laplace approximation, is
required for posterior inference in this case. We expect that the Laplace method
will produce worse approximations the more complex the target model. Since the
approximation is already unacceptable for our simple test model, we recommend
that Maud users use MCMC sampling in preference to Laplace approximation if
possible when fitting realistic Bayesian kinetic models.

@@

Graphical comparison of approximate posterior samples generated using Laplace
sampling (blue-green) with posterior samples generated using MCMC (dark
grey). (**Top left**) the two sets of samples clearly have different marginal
distributions for the overall log probability variable, indicating that the
Laplace samples do not accurately approximate the target distribution. (**Top
right**) The distribution of marginal posterior predictive log likelihood
values for training data flux measurements shows that the Laplace method
tended to yield much worse predictions compared with the true model (lower log
likelihood values indicate that the modelled and measured values are further
away). (**Bottom left**) The Laplace method also tended to produce worse flux
predictions for held-back test measurements. (**Bottom right**) The marginal
joint distribution of two parameters: $K_m^{A,rl}$ and $K_{cat}^{r1}$. The
Laplace method is not able to track the correct joint distribution for this
pair of parameters. This is unsurprising given that the target distribution
has position-dependent scales which are difficult for a linear approximation
to capture.

@@

## Effect of missing metabolite concentration measurements

@@

As might be expected, the model with missing measurements did not correctly
infer the missing ahcys concentrations (@fig-missing). Nonetheless, the
remaining measured metabolites were still well predicted, suggesting
that information about the network is still preserved despite the missing
measurements. Comparison of flux measurements in both models also indicated that
removing the ahcys measurement did not result in catastrophic model failure.

The missing measurements did affect Maud’s ability to infer parameter values
correctly (@fig-missing, lower left). The model with full metabolomics learned
the true value for the displayed dissociation constant, despite this value being
far from the mean of the corresponding marginal prior distribution. In contrast,
the model with missing measurements stayed in the neighbourhood of the prior.

@@

## Application to regulatory understanding

To demonstrate how Maud's output can be used to yield useful metabolic insights
we used the results of our case study to explain why the flux of the enzyme
GNMT is higher in dataset 1 than in dataset 12 (@fig-decomposition). GNMT
is an irreversible enzyme that is homotropically activated by its substrate,
competitively inhibited by its product and heterotropically inhibited by 5,10-
methylenetetrahydrofolate (mlthf). The complex regulation makes it the ideal
test case to elucidate regulatory changes.

Regulation can be separated into enzyme abundance, allostery and saturation,
and we can plot the marginal posterior distribution of the ratio of the
corresponding regulatory component in dataset 1 compared with dataset 12
(@fig-decomposition, right panel). A positive value indicates that the component
was increased in dataset 1 relative to dataset 12, with 0 indicating no
difference. The probability, according to our model, of the component acting in
each direction is given by the relative area under the curve on each side of the
zero point.

@@

Illustration of how analysing a system with Maud can yield actionable insights
about the underlying metabolic network.  (**Top Left**) A schematic of the
regulatory interactions associated with the enzyme $GNMT1$. Dashed green
lines represents allosteric activation, dashed red lines indicate allosteric
inhibition and solid red lines represent competitive inhibition. (**Bottom
Left**) Comparison of marginal posterior distributions for $GNMT$ flux in
datasets 1 and 2. (**Right**) Log-scale ratios of the regulatory elements
defined in \eqref{eq-decomposition}. Note that the $Reversibility$
and $k_cat$ components are excluded: this is because this reaction was modelled
as irreversible, and $k_{cat}$ was modelled as constant across datasets.
These plots identify why flux in dataset 12 is higher than in dataset 1: the
flux increase is due to allostery and saturation with no control from enzyme
concentration changes.

@@

As well as parameter values, Maud also performs inference for derived quantities
including the components of the regulatory decomposition described below in
\eqref{eq-decomposition}, log likelihoods, simulated measurement values
and metabolic control analysis coefficients [@kacser_control_1973].

@@

Maud’s kinetic model decomposes into factors:

@@

Metabolite and enzyme measurements are intended to represent the results of
quantitative metabolomics and proteomics experiments. The likelihood functions
for such measurements are:

@@

Both equations are log-normal generalised linear models with a standard
link function (the natural logarithm $\ln$) and known standard deviation
$ \sigma_{𝑐𝑜𝑛𝑐}$. The use of this measurement model is motivated by the
consideration that concentrations are constrained to be non-negative, so the
measurement model should avoid assigning positive probability mass to negative
metabolite concentration values. In addition, we expect the precision of most
metabolomics and proteomics experiments to be roughly proportional to the true
value of the measured quantity, which supports a measurement model with constant
coefficient of variation. The measurement standard deviations $\sigma_{𝑐𝑜𝑛𝑐}$
and $\sigma_{𝑒𝑛𝑧}$ are assumed to be known exactly for simplicity; plausible
values can be elicited by considering the likely coefficient of variation of the
measuring apparatus. Our measurement model improves on analyses of metabolomics
and proteomics data that assume a regression model with normally distributed
errors, whether explicitly using a standard linear model or implicitly using
ordinary least squares fitting.

Flux measurements, representing the results of quantitative fluxomics analyses,
are modelled using a likelihood function from a standard linear regression
model \eqref{eq-yflux}. Flux measurements can be obtained from isotope labelling
experiments using metabolic flux analysis, for example as described in (Young
2014). When entering flux measurements, it is important only to specify
measurements for a network’s free fluxes, as the values of some steady state
fluxes in a metabolic network are constrained by others, with the result that
dependent fluxes cannot typically be measured separately. If measurements of
multiple dependent fluxes are entered, information will inappropriately be
double counted.

@@

@@

Maud's parameters are connected with measurable quantities via the steady state
equation \eqref{eq-3}. Posterior sampling using adaptive Hamiltonian Monte
Carlo requires repeatedly solving \eqref{eq-3} and finding the gradients of its
solution with respect to all model parameters. This must be done numerically as
analytic solutions are not available for realistic kinetic models.

@@

To solve \eqref{eq-3} and find its gradients, Maud uses a hybrid
method involving two numerical solvers from the SUNDIALS suite
[@serbanCVODESSensitivityEnabledODE2005]: CVODES and IDAS via their
interface from Stan. The hybrid method follows that proposed by
@margossianComputingSteadyStates2018, and involves numerically evolving the ODE
system for a short period of time, then using the difference between the evolved
and starting concentrations as the target for a numerical algebra solver.

@@

This research was funded by the Novo Nordisk Foundation (Grant numbers
NNF20CC0035580 NNF14OC0009473).
